---
title: "Explode, edit, compare, render"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explode, edit, compare, render}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
have_dragmapr <- requireNamespace("dragmapr", quietly = TRUE) &&
  "d_state" %in% getNamespaceExports("dragmapr")
```

The full `explodemap` loop is **explode, measure, edit, compare, render**:

```
explode_grouped()  ->  displacement_magnitudes()  ->  d_edit()  ->  compare_layouts()  ->  as_sf()
   (compute)              (measure)                   (compose)        (compare)             (export)
```

This vignette walks the whole loop on a small synthetic grid. The editorial
pass is shown two ways: interactively in the dragmapr editor, and as a
scripted nudge in pure R (handy for reproducible tweaks and for CI, where no
browser is available).

## 1. Explode

```{r explode}
library(explodemap)

unit_square <- function(x0, y0, size = 40000) {
  sf::st_polygon(list(rbind(
    c(x0, y0), c(x0 + size, y0), c(x0 + size, y0 + size),
    c(x0, y0 + size), c(x0, y0)
  )))
}

grid <- expand.grid(col = 0:2, row = 0:2)
regions <- sf::st_sf(
  region = rep(c("A", "B", "C"), each = 3),
  unit   = sprintf("u%02d", seq_len(9)),
  geometry = sf::st_sfc(
    lapply(seq_len(9), function(i) unit_square(grid$col[i] * 50000, grid$row[i] * 50000)),
    crs = 3857
  )
)

layout <- explode_grouped(regions, region_col = "region", plot = FALSE, quiet = TRUE)
plot(layout)
```

## 2. Measure the layout

Before editing, quantify what the algorithm did. `displacement_magnitudes()`
gives the per-feature movement in metres:

```{r measure}
dm <- displacement_magnitudes(layout)
tapply(dm$distance, as_sf(layout)$region, function(d) round(c(mean = mean(d), max = max(d))))
```

## 3. Hand off to dragmapr

`as_dragmapr_state()` converts the computed anchors into a reusable
`dragmapr_state` — the composition object the editor and the renderers share:

```{r handoff, eval = have_dragmapr}
state <- as_dragmapr_state(layout)
state
```

## 4. Edit

Open the editor seeded with the computed layout. In Shiny, capture each edit
back into a state with `dragmapr::d_widget_state()`:

```{r edit-interactive, eval = FALSE}
dragmapr::d_edit(layout, state = state)
```

For a reproducible tweak — or anywhere without a browser — apply the same
kind of edit as data. Here region "B" is nudged 50 km east:

```{r edit-scripted}
nudge <- data.frame(
  region = "B",
  dx_m = 50000,
  dy_m = 0
)
edited <- update_exploded_layout(layout, nudge, update_plots = FALSE)
```

## 5. Compare before and after

`compare_layouts()` summarises how the edit changed per-feature displacement:

```{r compare}
cmp <- compare_layouts(layout, edited)
cmp$summary
head(cmp$per_feature)
```

## 6. Render and export

Render the edited composition statically (writes a PNG next to this
document's working directory when run interactively):

```{r render, eval = FALSE}
dragmapr::render_dragged_map(
  as_sf(edited, which = "grouped"),
  region_col = "region",
  state = state,
  file = "edited.png"
)
```

And convert the final layout back to plain `sf` for downstream GIS work —
file export, spatial joins, or further `sf` analysis:

```{r export}
final_sf <- as_sf(edited)
sf::st_write(final_sf, file.path(tempdir(), "edited.gpkg"), quiet = TRUE)
list.files(tempdir(), pattern = "edited")
```

The loop closes: `final_sf` is an ordinary `sf` data frame, so anything that
reads simple features can pick up where the editor left off.
