---
title: "Getting started with explodemap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with explodemap}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`explodemap` generates hierarchical exploded-view maps for dense
administrative boundary data. It applies rigid-body translations to
polygon geometries so that features are visually separated while each
feature's internal geometry is preserved exactly.

The basic two-level workflow is:

1. Group units into regions using a column in your data.
2. Displace units within and across those regions using a centroid-driven
   vector field with analytically derived parameters.

For the two-level core described in the paper, the package is designed
to satisfy three key properties:

- **Proposition 1:** each polygon's shape, area, perimeter, and topology
  are preserved exactly.
- **Proposition 2:** the radial ordering of units within each region is
  preserved.
- **Proposition 3:** no unit is displaced by more than
  `alpha_r + alpha_l` metres.

## Input requirements

`explodemap` expects an `sf` object with:

- Polygon or multipolygon geometries
- A grouping column identifying the parent region for each unit
- A **projected** CRS with linear metre units

Geographic coordinates such as EPSG:4326 (longitude/latitude) must be
transformed before use. For U.S. work, a state plane, UTM, or Albers-type
projected CRS is usually appropriate.

```{r setup}
library(sf)
library(explodemap)
```

## A minimal example

We create a small synthetic dataset with four square polygons in two
regions.

```{r synthetic-data}
sq <- function(xmin, ymin, size = 1000) {
  st_polygon(list(matrix(
    c(xmin, ymin,
      xmin + size, ymin,
      xmin + size, ymin + size,
      xmin, ymin + size,
      xmin, ymin),
    ncol = 2,
    byrow = TRUE
  )))
}

geom <- st_sfc(
  sq(0, 0), sq(3000, 0),       # Region A
  sq(12000, 0), sq(15000, 0),  # Region B
  crs = 3857
)

x <- st_sf(
  id     = c("a1", "a2", "b1", "b2"),
  region = c("A", "A", "B", "B"),
  geometry = geom
)

x
```

## Running the explosion

The simplest entry point is `explode_sf()`. Pass your sf object and the
name of the grouping column:

```{r explode-basic}
result <- explode_sf(x, region_col = "region", plot = FALSE)
```

The returned object is an S3 object of class `exploded_map`:

```{r class-check}
class(result)
names(result)
```

It contains the original and exploded geometries, a WGS84 export-ready
version, derived statistics and parameters, plots, and diagnostics.

## Diagnostics

`print()` shows geometry statistics and derived parameters:

```{r print}
print(result)
```

`summary()` adds implied gamma coefficients that are useful for
calibration work:

```{r summary}
summary(result)
```

## Plotting

```{r plot-exploded}
plot(result)
```

You can also view both original and exploded layouts side by side:

```{r plot-both, fig.width = 10}
plot(result, "both")
```

## Calibration output

`calibration_row()` returns a one-row data frame suitable for combining
across datasets when building a cross-dataset calibration table:

```{r calibration}
calibration_row(result)
```

## Manual parameter overrides

By default, `explodemap` derives displacement parameters analytically
from dataset geometry using the paper's two closed-form results. You can
also supply parameters directly. Overrides may be supplied
independently, so you can tune regional separation without changing
local expansion, or vice versa:

```{r manual}
manual <- explode_sf(
  x,
  region_col = "region",
  alpha_r = 100,
  alpha_l = 200,
  plot = FALSE
)

manual$params
```

```{r partial-manual}
more_region_gap <- explode_sf(
  x,
  region_col = "region",
  alpha_r = result$params$alpha_r * 1.5,
  plot = FALSE
)

more_region_gap$params
```

## Optional collision refinement

The two-level algorithm is the clean paper model. For dense municipal
cores, you can add a bounded refinement pass that nudges close
same-region neighbors apart after the analytical displacement:

```{r refine-example, eval=FALSE}
refined <- explode_sf(
  x,
  region_col = "region",
  refine = TRUE,
  refine_min_gap = 0.15,
  refine_max_shift = 0.10,
  plot = FALSE
)

refined$refinement
```

```{r refine, echo=FALSE}
has_refinement <- "refine" %in% names(formals(explode_sf))

if (has_refinement) {
  refined <- explode_sf(
    x,
    region_col = "region",
    refine = TRUE,
    refine_min_gap = 0.15,
    refine_max_shift = 0.10,
    plot = FALSE
  )

  refined$refinement
} else {
  message("This installed explodemap build does not include refinement yet.")
}
```

`refine_max_shift` caps the extra correction per feature, so the
refinement remains a small display adjustment rather than a replacement
for the displacement model. Use `refine_within = "all"` when the
remaining crowding crosses region boundaries.

## Centroid options

For irregular or multipart polygons, `"point_on_surface"` may be
preferable to the default geometric centroid:

```{r centroid-mode}
pos <- explode_sf(
  x,
  region_col = "region",
  centroid_fun = "point_on_surface",
  plot = FALSE
)
```

## Using TIGER/Line data for U.S. states

`explode_state()` downloads U.S. Census TIGER/Line boundaries
automatically and groups municipalities by a county-to-region mapping:

```{r tiger-example, eval = FALSE}
nj <- explode_state(
  state_fips = "34", crs = 32118,
  region_map = list(
    North   = c("Bergen", "Essex", "Hudson", "Morris",
                "Passaic", "Sussex", "Union", "Warren"),
    Central = c("Hunterdon", "Mercer", "Middlesex",
                "Monmouth", "Somerset"),
    South   = c("Atlantic", "Burlington", "Camden", "Cape May",
                "Cumberland", "Gloucester", "Ocean", "Salem")
  ),
  label = "New Jersey"
)
```

Downloaded data is cached locally so subsequent runs are faster.

## Using a lookup table

`explode_sf_with_lookup()` joins an external lookup table to your sf
object before exploding:

```{r lookup-example, eval = FALSE}
groups <- read.csv("region_assignments.csv")

result <- explode_sf_with_lookup(
  my_sf,
  join_col   = "GEOID",
  lookup     = groups,
  lookup_key = "geoid",
  region_col = "region"
)
```

Unmatched units are labelled `"Other"`. This is useful when a lookup
table is incomplete. You can include or exclude those units using
`allow_other`.

## Export

The `export` parameter supports three modes:

- `NULL` (default): no export
- `TRUE`: auto-named GeoJSON file
- A file path string: explicit output location

```{r export-example, eval = FALSE}
result <- explode_sf(
  my_sf,
  region_col = "region",
  export = "output.geojson"
)
```

## Notes

- Always use a projected CRS before running the algorithm.
- The two-level core preserves each feature's internal geometry exactly
  through rigid-body translation.
- Parameter derivation is deterministic and reproducible: the same
  dataset and gamma coefficients always produce the same output.

## Next steps

See `vignette("grouped-layouts")` for the three-level extension using
`explode_grouped()`, which adds region-block anchor placement for larger
multi-region or national layouts.
