---
title: "Grouped layouts and anchor placement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped layouts and anchor placement}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`explode_grouped()` extends the core two-level exploded-map workflow
with a three-level hierarchy for multi-region or national-scale layouts.

This grouped extension is most useful when a standard two-level
explosion still leaves region blocks visually crowded or difficult to
compare across a larger spatial extent.

The three levels are:

1. **Level 1 — Local explosion.** Units within each parent region are
   displaced using the centroid-driven field from the core algorithm.
   The geometric guarantees of Propositions 1--3 apply at this level.

2. **Level 2 — Radial anchor placement.** Parent region centroids are
   displaced radially outward from the national centroid to generate
   initial anchor positions for each region block.

3. **Level 3 — Collision-aware refinement** (optional). Overlapping
   anchors are iteratively repelled while a spring term maintains
   proximity to the original radial targets.

Levels 2 and 3 preserve structural grouping and directional
correspondence rather than topological coverage. The formal geometric
guarantees of Propositions 1--3 apply strictly at Level 1.

```{r setup}
library(sf)
library(explodemap)
```

## A synthetic grouped example

We create a small dataset with six units in three regions, spread across
a wider spatial extent than the two-region example in
`vignette("getting-started")`.

```{r synthetic-grouped}
sq <- function(xmin, ymin, size = 1000) {
  st_polygon(list(matrix(
    c(xmin, ymin,
      xmin + size, ymin,
      xmin + size, ymin + size,
      xmin, ymin + size,
      xmin, ymin),
    ncol = 2,
    byrow = TRUE
  )))
}

geom <- st_sfc(
  sq(0, 0), sq(3000, 0),       # R1
  sq(9000, 0), sq(12000, 0),   # R2
  sq(24000, 0), sq(27000, 0),  # R3
  crs = 3857
)

x <- st_sf(
  id     = paste0("u", 1:6),
  region = c("R1", "R1", "R2", "R2", "R3", "R3"),
  geometry = geom
)
```

## Anchor modes

`explode_grouped()` supports three anchor placement modes:

| Mode | Description |
|------|-------------|
| `"auto"` | Level 2 only — radial placement without collision resolution |
| `"auto_collision"` | Level 2 + Level 3 — radial placement with iterative refinement |
| `"manual"` | User-supplied anchor coordinates |

## Automatic grouped layout

The simplest call uses `mode = "auto"`:

```{r auto-mode}
g_auto <- explode_grouped(
  x,
  region_col = "region",
  mode       = "auto",
  plot       = FALSE
)

class(g_auto)
print(g_auto)
```

The result is a `grouped_exploded_map` S3 object, which also inherits
from `exploded_map`:

```{r auto-names}
names(g_auto)
```

Plot the grouped layout:

```{r auto-plot}
plot(g_auto)
```

## Collision-aware grouped layout

`mode = "auto_collision"` adds Level 3 refinement. Overlapping region
blocks are iteratively repelled while a spring term pulls them back
toward their radial targets:

```{r collision-mode}
g_collide <- explode_grouped(
  x,
  region_col = "region",
  mode       = "auto_collision",
  plot       = FALSE
)

plot(g_collide)
```

The anchor table reports the resulting block radii and coordinates:

```{r convergence}
g_collide$anchors[, c("region", "block_radius", "anchor_x", "anchor_y")]
```

## Viewing all stages

`plot(g, "all")` shows the original, locally exploded, and grouped
layouts side by side:

```{r all-views, fig.width = 10, fig.height = 4}
plot(g_collide, "all")
```

## Inspecting anchor positions

`layout_regions()` computes anchor positions as a standalone step, which
is useful for custom workflows or manual adjustment:

```{r layout-regions}
anchors <- layout_regions(
  x,
  region_col = "region",
  mode       = "auto"
)

anchors[, c("region", "block_radius", "n_units", "anchor_x", "anchor_y")]
```

## Manual anchors

You can edit anchor positions and pass them back into
`explode_grouped()`:

```{r manual-anchors}
manual_anchors <- anchors
manual_anchors$anchor_x <- manual_anchors$anchor_x + c(0, 500, 1000)
manual_anchors$anchor_y <- manual_anchors$anchor_y + c(0, 250, 500)

g_manual <- explode_grouped(
  x,
  region_col = "region",
  mode       = "manual",
  anchors    = manual_anchors,
  plot       = FALSE
)

plot(g_manual)
```

## Key parameters

Level 1 parameters control local displacement within regions:

| Parameter | Default | Purpose |
|-----------|---------|---------|
| `alpha_l` | derived | Local expansion magnitude (metres) |
| `p` | 1.25 | Distance-scaling exponent |
| `gamma_l` | 1.136 | Local clearance coefficient (used if `alpha_l` is NULL) |

Level 2 and Level 3 parameters control anchor placement and refinement:

| Parameter | Default | Purpose |
|-----------|---------|---------|
| `kappa` | 1.8 | Radial expansion factor |
| `padding` | 50000 | Base padding (map units) |
| `delta` | 15000 | Log-density scaling factor |
| `lambda` | 0.18 | Spring coefficient for refinement |
| `eta` | 0.18 | Repulsion step size for refinement |
| `padding_sep` | 20000 | Minimum separation between blocks |

For real-world data, these defaults are tuned for national-scale U.S.
layouts such as states grouped into larger reporting regions. For
smaller or larger extents, adjust `padding`, `delta`, and
`padding_sep` to match your coordinate system and visual scale.

## Summary

```{r summary}
summary(g_collide)
```

## What the grouped object stores

| Field | Contents |
|-------|----------|
| `sf_orig` | Original input geometries |
| `sf_local` | Geometries after Level 1 local explosion |
| `sf_grouped` | Final geometries after anchor displacement |
| `sf_grouped_wgs` | Final geometries in WGS84 (EPSG:4326) |
| `stats` | Geometry statistics from `compute_stats()` |
| `params` | All parameters used |
| `anchors` | Anchor table with block radii and positions |
| `plots` | ggplot objects for original, local, and grouped layouts |
| `diagnostics` | Label, region column, centroid mode, and anchor mode |

## Scope of guarantees

The local explosion stage (Level 1) preserves the package's core
geometric guarantees: exact feature-level geometry preservation
(Proposition 1), radial ordering within regions (Proposition 2), and
bounded displacement (Proposition 3).

The grouped anchor stage (Levels 2--3) is a higher-level layout
extension. It preserves grouping structure and directional
correspondence rather than topological coverage.

## Next steps

For the core two-level workflow and parameter derivation, see
`vignette("getting-started")`.

For full paper-scale examples including cross-state calibration, Canada,
and HHS grouped layouts, see
`vignette("reproducing-paper-examples")`.
